---
title: 'Final Report'
author: "Iman Byndloss, Leah Li, and Rachel Williams" 
date: "12/09/2024"
output: pdf_document
toc: true
---
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
setwd("/home/guest/ByndlossLiWilliams")

# Load your packages
library(tidyverse)
library(lubridate)
library(dplyr)
library(here)
library(ggplot2)
library(knitr)

here()

# Set your ggplot theme
my_theme <- 

# Load your datasets
Daily_Species_Counts <- read.csv(
  file = here("./Data/Processed/Daily_Species_Counts.csv"),
  stringsAsFactors = TRUE) 

Monthly_Species_Counts <- read.csv(
  file = here("./Data/Processed/Monthly_Species_Counts.csv"),
  stringsAsFactors = TRUE) 

Daily_Merge <- read.csv(
  file = here("./Data/Processed/Daily_Merge.csv"),
  stringsAsFactors = TRUE) 

Monthly_Merge <- read.csv(
  file = here("./Data/Processed/Monthly_Merge.csv"),
  stringsAsFactors = TRUE) 
```

```


# Rationale and Research Questions

The upcoming analyses seek to answer the following research questions: Does the overall activity (daily and monthly) for mammalian species differ across three Davidson County parks (Beaman, Bells Bend, and Mill Creek)? And do any park attributes (i.e., classification, size and disruptive activities) have a significant impact on the activity of the two most common species? Assuming null hypotheses, there is no significant difference between overall daily and monthly activity for mammals in the three parks, and none of the tested park attributes significantly impact the activity of eastern gray squirrels and white-tailed deer. 

To describe the motivation behind this research, we are interested in understanding how different factors influence mammalian species activity in urban parks. Parks like Beaman, Bells Bend, and Mill Creek not only provide essential green spaces for humans but also serve as critical habitats for wildlife. Our curiosity lies in exploring how park attributes, such as size and unique features, shape species behavior over time. By understanding these dynamics, we hope to foster greater appreciation for biodiversity in urban environments and support park management practices. 


\newpage

# Dataset Information

The datasets we used were collected from wildlife insights and the Nashville open data portal (https://data.nashville.gov/). The wildlife insights data was the camera trap data collected for each park. It is from Professor Malu Jorge's lab at Vanderbilt University. This data contains deployments from 2019-2022, but we focused on the deployments from 2021-2022 because the target parks 

```{r}
names(Daily_Species_Counts)
summary(Daily_Species_Counts)
```

\newpage

# Exploratory Analysis 

The reason we picked the target parks is...


To answer our second question, we needed to know which species were most abundant at the target parks. To do this, we aggregated the data to add up the adjusted counts for each species. Once we did this, we sorted it in descending order and looked at which species were the two most abundant.

```{r}
aggregated_data <- Daily_Species_Counts %>%
  group_by(Species) %>%
  summarise(total_count = sum(CountAdjusted, na.rm = TRUE))
  
sorted_data <- aggregated_data %>%
  arrange(desc(total_count))
  
most_abundant_species <- head(sorted_data, 5)

kable(most_abundant_species, caption = "Most Abundant Species")
```
According to these results, White-tailed deer and Eastern Gray Squirrels were the top two most abundant species. 

\newpage

# Analysis

<Question 1: The first question we wanted to answer is whether daily and monthly species activity differs between the target parks. We looked at Beaman, Bell's Bend, and Mill's Creek Park.>

Does species' daily and monthly activity differ between parks?  

#Daily Analysis

```{r}
# Perform the two-way ANOVA with interaction
daily_anova_result <- aov(CountAdjusted ~ ParkName * Species, data = Daily_Species_Counts)

# View the ANOVA summary
summary(daily_anova_result)
```

```{r}
# If the interaction term is significant, perform a post-hoc test
Daily_Species_Counts$ParkName <- as.factor(Daily_Species_Counts$ParkName)
Daily_Species_Counts$Species <- as.factor(Daily_Species_Counts$Species)

# Tukey's HSD test for the interaction term (ParkName * Species)
library(agricolae)
daily_tukey_result <- TukeyHSD(daily_anova_result, "ParkName:Species")

# Extract significant results
daily_tukey_df <- as.data.frame(daily_tukey_result$`ParkName:Species`)
daily_tukey_df$Comparison <- rownames(daily_tukey_result$`ParkName:Species`)
rownames(daily_tukey_df) <- NULL

# Filter significant comparisons
daily_significant_tukey <- subset(daily_tukey_df, `p adj` < 0.05)

# Print the top 10 significant results
head(daily_significant_tukey, 10)
```

#Monthly Analysis

```{r}
# Perform the two-way ANOVA with interaction
monthly_anova_result <- aov(CountAdjusted ~ ParkName * Species, data = Monthly_Species_Counts)

# View the ANOVA summary
summary(monthly_anova_result)
```

```{r}
# If the interaction term is significant, perform a post-hoc test
Monthly_Species_Counts$ParkName <- as.factor(Monthly_Species_Counts$ParkName)
Monthly_Species_Counts$Species <- as.factor(Monthly_Species_Counts$Species)

# Tukey's HSD test for the interaction term (ParkName * Species)
library(agricolae)
monthly_tukey_result <- TukeyHSD(monthly_anova_result, "ParkName:Species")

# Extract significant results
monthly_tukey_df <- as.data.frame(monthly_tukey_result$`ParkName:Species`)
monthly_tukey_df$Comparison <- rownames(monthly_tukey_result$`ParkName:Species`)
rownames(monthly_tukey_df) <- NULL

# Filter significant comparisons
monthly_significant_tukey <- subset(monthly_tukey_df, `p adj` < 0.05)

# Print the top 10 significant results
head(monthly_significant_tukey, 10)
```

Question 2: The second question we wanted to answer is what factors might be causing different activity levels between parks. This study focused on the size, classification, and disruptive activities of each park. We used Deer and squirrels for this analysis because they were the most abundant species at all three parks. 

What factors might be causing different activity levels between parks for WHite-tailed Deer and Grey Squirrels? Does classification, size, or disruptive activities at each park cause changes in activity levels?

#Creating Daily and Monthly merged sets for Deer and Squirrels
```{r}
dmSquirrel <- Daily_Merge %>% 
  filter(Species == "Eastern Gray Squirrel")

dmDeer <- Daily_Merge %>% 
  filter(Species == "White-tailed Deer") 
  
mmSquirrel <- Monthly_Merge %>%
  filter(Species == "Eastern Gray Squirrel")
  
mmDeer <- Monthly_Merge %>%
  filter(Species == "White-tailed Deer")
```

#Daily t-tests
```{r}
#Classification t-tests: 
SQ_t_test_class <- t.test(CountAdjusted ~ Classification, data = dmSquirrel)
SQ_t_test_class

D_t_test_class <- t.test(CountAdjusted ~ Classification, data = dmDeer)
D_t_test_class

#Park Size t-tests
SQ_t_test_size <- t.test(CountAdjusted ~ Size, data = dmSquirrel)
SQ_t_test_size

D_t_test_size <- t.test(CountAdjusted ~ Size, data = dmDeer)
D_t_test_size

#Disruption t-tests
SQ_t_test_disrupt <- t.test(CountAdjusted ~ DisruptiveActivities, data = dmSquirrel)
SQ_t_test_disrupt

D_t_test_disrupt <- t.test(CountAdjusted ~ DisruptiveActivities, data = dmDeer)
D_t_test_disrupt
```

#Monthly t-tests
```{r}
#Classification t-test
SQ_t_test_class2 <- t.test(CountAdjusted ~ Classification, data = mmSquirrel)
SQ_t_test_class2

D_t_test_class2 <- t.test(CountAdjusted ~ Classification, data = mmDeer)
D_t_test_class2

#Park size t-tests
SQ_t_test_size2 <- t.test(CountAdjusted ~ Size, data = mmSquirrel)
SQ_t_test_size2

D_t_test_size2 <- t.test(CountAdjusted ~ Size, data = mmDeer)
D_t_test_size2

#Disruption t-tests
SQ_t_test_disrupt2 <- t.test(CountAdjusted ~ DisruptiveActivities, data = mmSquirrel)
SQ_t_test_disrupt2

D_t_test_disrupt2 <- t.test(CountAdjusted ~ DisruptiveActivities, data = mmDeer)
D_t_test_disrupt2
```

\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section>