---
title: "Final Project: Data Analysis"
author: "Iman Byndloss, Leah Li, and Rachel Williams"
output: html_document
date: "2024-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Set Up; Uploading Datasets

```{r,  set up R session, message = FALSE, warning=FALSE}
#Load packages
library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)

#Upload species dataset
Daily_Species_Count.data <- read.csv(
  file = here("./Data/Processed/Daily_Species_Counts.csv"),
  stringsAsFactors = TRUE) 
```

#Two-way ANOVA Test

```{r, First ANOVA, message = FALSE, warning=FALSE}
# Perform the two-way ANOVA with interaction
first_anova_result <- aov(CountAdjusted ~ ParkName * Species, data = Daily_Species_Count.data)

# View the ANOVA summary
summary(first_anova_result)


```

>ANOVA Result Interpretation:
>The two-way ANOVA analysis here evaluated the effects of park location and species type, as well as their interaction, on species activity levels (measured as adjusted counts). The results revealed a significant main effect of park location (F=99.91, p<0.001), indicating that species activity levels varied significantly between parks. Similarly, a highly significant main effect was observed for species type (F=142.86, p<0.001), suggesting that the activity levels differed substantially across the species studied.

>In addition to the main effects, the interaction between park location and species type was also significant (F=127.19, p<0.001). This finding highlights that the effect of park location on species activity is not consistent across all species. In other words, the differences in species activity levels between parks are influenced by the specific species being observed.

>These results suggest that both park-specific factors (e.g., habitat type, management practices) and species-specific traits (e.g., behavior, ecological preferences) play critical roles in determining activity levels. Moreover, the significant interaction effect emphasizes the importance of considering the combined influence of these two factors when analyzing species activity patterns.

## Post-Hoc Test

```{r, Post-Hoc Test after ANOVA, message = FALSE, warning=FALSE}
# If the interaction term is significant, perform a post-hoc test
Daily_Species_Count.data$ParkName <- as.factor(Daily_Species_Count.data$ParkName)
Daily_Species_Count.data$Species <- as.factor(Daily_Species_Count.data$Species)

# Tukey's HSD test for the interaction term (ParkName * Species)
library(agricolae)
tukey_result <- TukeyHSD(first_anova_result, "ParkName:Species")

# Extract significant results
tukey_df <- as.data.frame(tukey_result$`ParkName:Species`)
tukey_df$Comparison <- rownames(tukey_result$`ParkName:Species`)
rownames(tukey_df) <- NULL

# Filter significant comparisons
significant_tukey <- subset(tukey_df, `p adj` < 0.05)

# Print the top 10 significant results
head(significant_tukey, 10)

```

>Post-Hoc Result Interpretation

>1. Species Differences Between Parks: The activity of "Nine-banded Armadillos" exhibited notable variation between parks.  For instance, their activity was significantly lower in Bells Bend Park compared to Mill Creek Park # (diff = -1.53, 95% CI [-1.71, -1.35], p adj < 0.001). Similarly, their activity in Alvin G. Beaman Park was also significantly lower compared to Mill Creek Park  (diff = -1.81, 95% CI [-1.99, -1.63], p adj < 0.001). These results suggest that the habitat or environmental factors in Mill Creek Park are more conducive to higher activity levels for this species.

>2.Within-Park Differences Between Species: In Mill Creek Park, "Muskrat" activity was significantly lower than that of "Eastern Gray Squirrels" (diff = -1.81, 95% CI [-1.99, -1.63], p adj < 0.001). This could indicate competitive dynamics, differences in habitat preference,  or time-specific activity patterns between these species.

>3. Activity Peaks for White-Tailed Deer: The White-tailed Deer in Bells Bend Park showed significantly higher activity compared to other species within the park (diff = 1.15, 95% CI [0.97, 1.32], p adj < 0.001). This finding suggests that this species might be particularly dominant in this park, potentially due to favorable environmental or ecological conditions.

>4. Notable Inter-Park Variation in Woodchuck Activity: The activity of "Woodchucks" was significantly lower in Alvin G. Beaman Park compared to Bells Bend Park (diff = -1.33, 95% CI [-1.51, -1.15], p adj < 0.001). This highlights how habitat differences between parks may influence the presence or detectability of certain species.  Interpretation: These results reveal substantial variation in species activity both within and between parks. Certain species, such as "Nine-banded Armadillos" and "White-tailed Deer," exhibit pronounced activity trends influenced by the park's environmental conditions. The findings emphasize the importance of localized habitat characteristics and interspecies dynamics in shaping activity patterns.

