---
title: "Final Project: Data Analysis"
author: "Iman Byndloss, Leah Li, and Rachel Williams"
output: html_document
date: "2024-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The upcoming analyses seek to answer the following research questions: Do overall daily and monthly mammalian species activity differ across three Nashville parks (Beaman, Bells Bend, and Mill Creek)? And do any park attributes (i.e., classification, size and disruptive activities) have a significant impact on the activity of three potentially important species? Assuming null hypotheses, there is no significant difference between overall daily and monthly activity for mammals in the three parks, and none of the tested park attributes significantly impact the activity of eastern gray squirrels, white-tailed deer, and coyote. These assumptions will be tested through two-way anova (for overall activity across parks) and two-sample t-tests (for the impact of park attributes on specific activity).

To describe the motivation behind this research, we are interested in understanding how different factors influence mammalian species activity in urban parks. Parks like Beaman, Bells Bend, and Mill Creek not only provide essential green spaces for humans but also serve as critical habitats for wildlife. Our curiosity lies in exploring how park attributes, such as size and unique features, shape species behavior over time. By understanding these dynamics, we hope to foster greater appreciation for biodiversity in urban environments and support park management practices. 


## Set Up (loading packages and uploading datasets)

```{r,  set up R session, message = FALSE, warning=FALSE}
# Load relevant packages
library(tidyverse)
library(lubridate)
library(dplyr)
library(here)
library(ggplot2)

# Load relevant processed data
Daily_Species_Counts <- read.csv(
  file = here("./Data/Processed/Daily_Species_Counts.csv"),
  stringsAsFactors = TRUE) 

Monthly_Species_Counts <- read.csv(
  file = here("./Data/Processed/Monthly_Species_Counts.csv"),
  stringsAsFactors = TRUE) 

Daily_Merge <- read.csv(
  file = here("./Data/Processed/Daily_Merge.csv"),
  stringsAsFactors = TRUE) 

Monthly_Merge <- read.csv(
  file = here("./Data/Processed/Monthly_Merge.csv"),
  stringsAsFactors = TRUE) 
```

# Two-way ANOVA Test 

## Daily Species Activity (Park and Species)

```{r, First ANOVA, message = FALSE, warning=FALSE}
# Perform the two-way ANOVA with interaction
daily_anova_result <- aov(CountAdjusted ~ ParkName * Species, data = Daily_Species_Counts)

# View the ANOVA summary
summary(daily_anova_result)
```

> ANOVA Result Interpretation:

> The two-way ANOVA analysis here evaluated the effects of park and species, as well as their interaction, on daily activity levels (measured as adjusted counts). In terms of the main effects, the results revealed daily activity levels differ significantly across parks (Df = 2, F-value = 99.91, p-value = <2e-16) and species (Df = 15, F-value = 142.86, p-value = <2e-16). In terms of the interaction effect, the interaction between park and species was also significant (Df = 19, F-value = 127.19, p-value = <2e-16). This finding highlights that the effect of park on daily activity is not consistent across all species and the effect of species on daily activity is not consistent across all parks. 

> These results suggest that both park-specific factors (e.g., habitat type, management practices) and species-specific traits (e.g., behavior, ecological preferences) play critical roles in determining daily activity levels. Moreover, the significant interaction effect emphasizes the importance of considering the combined influence of these two factors when analyzing daily activity.

### Post-Hoc Test

```{r, Post-Hoc Test after ANOVA, message = FALSE, warning=FALSE}
# If the interaction term is significant, perform a post-hoc test
Daily_Species_Counts$ParkName <- as.factor(Daily_Species_Counts$ParkName)
Daily_Species_Counts$Species <- as.factor(Daily_Species_Counts$Species)

# Tukey's HSD test for the interaction term (ParkName * Species)
library(agricolae)
daily_tukey_result <- TukeyHSD(daily_anova_result, "ParkName:Species")

# Extract significant results
daily_tukey_df <- as.data.frame(daily_tukey_result$`ParkName:Species`)
daily_tukey_df$Comparison <- rownames(daily_tukey_result$`ParkName:Species`)
rownames(daily_tukey_df) <- NULL

# Filter significant comparisons
daily_significant_tukey <- subset(daily_tukey_df, `p adj` < 0.05)

# Print the top 10 significant results
head(daily_significant_tukey, 10)
```

>Post-Hoc Result Interpretation:

> In total,  the post-hoc test for daily activity levels revealed 214 significant comparisons.

> The top 10 most significant results are between 1) Beaman Park eastern gray squirrel and Bells Bend Park American beaver, 2) Bells Bend Park eastern gray squirrel and Bells Bend Park American beaver, 3) Mill Creek Park eastern gray squirrel and Bells Bend Park American beaver, 4) Bells Bend Park nine-banded armadillo and Bells Bend Park American beaver, 5) Bells Bend Park northern raccoon and Bells Bend Park American beaver, 6) Mill Creek Park northern raccon and Bells Bend Park American beaver, 7) Beaman Park white-tailed deer and Bells Bend Park American beaver, 8) Bells Bend Park white-tailed deer and Bells Bend Park American beaver, 9) Beaman Park eastern gray squirrel and Mill Creek Park American beaver, and 10) Bells Bend eastern gray squirrel and Mill Creek Park American beaver. 

>1. Species Differences Between Parks: The activity of "Nine-banded Armadillos" exhibited notable variation between parks.  For instance, their activity was significantly lower in Bells Bend Park compared to Mill Creek Park # (diff = -1.53, 95% CI [-1.71, -1.35], p adj < 0.001). Similarly, their activity in Alvin G. Beaman Park was also significantly lower compared to Mill Creek Park  (diff = -1.81, 95% CI [-1.99, -1.63], p adj < 0.001). These results suggest that the habitat or environmental factors in Mill Creek Park are more conducive to higher activity levels for this species.

>2.Within-Park Differences Between Species: In Mill Creek Park, "Muskrat" activity was significantly lower than that of "Eastern Gray Squirrels" (diff = -1.81, 95% CI [-1.99, -1.63], p adj < 0.001). This could indicate competitive dynamics, differences in habitat preference,  or time-specific activity patterns between these species.

>3. Activity Peaks for White-Tailed Deer: The White-tailed Deer in Bells Bend Park showed significantly higher activity compared to other species within the park (diff = 1.15, 95% CI [0.97, 1.32], p adj < 0.001). This finding suggests that this species might be particularly dominant in this park, potentially due to favorable environmental or ecological conditions.

>4. Notable Inter-Park Variation in Woodchuck Activity: The activity of "Woodchucks" was significantly lower in Alvin G. Beaman Park compared to Bells Bend Park (diff = -1.33, 95% CI [-1.51, -1.15], p adj < 0.001). This highlights how habitat differences between parks may influence the presence or detectability of certain species.  Interpretation: These results reveal substantial variation in species activity both within and between parks. Certain species, such as "Nine-banded Armadillos" and "White-tailed Deer," exhibit pronounced activity trends influenced by the park's environmental conditions. The findings emphasize the importance of localized habitat characteristics and interspecies dynamics in shaping activity patterns.

## Monthly Species Activity (Park and Species)

```{r, Second ANOVA, message = FALSE, warning=FALSE}
# Perform the two-way ANOVA with interaction
monthly_anova_result <- aov(CountAdjusted ~ ParkName * Species, data = Monthly_Species_Counts)

# View the ANOVA summary
summary(monthly_anova_result)
```

> ANOVA Result Interpretation:

> The two-way ANOVA analysis here evaluated the effects of park and species, as well as their interaction, on monthly activity levels (measured as adjusted counts). In terms of the main effects, the results revealed daily activity levels differ significantly across parks (Df = 2, F-value = 11.81, p-value = 1.11e-05) and species (Df = 15, F-value = 16.88, p-value = <2e-16). In terms of the interaction effect, the interaction between park and species was also significant (Df = 19, F-value = 15.03, p-value = <2e-16). 

> With these results being similar to the results for daily activity, it can be understood that the effects of park-specific, species-specific, and interactive factors are consistent over longer time scales. 

### Post-Hoc Test

```{r, Post-Hoc Test after second ANOVA, message = FALSE, warning=FALSE}
# If the interaction term is significant, perform a post-hoc test
Monthly_Species_Counts$ParkName <- as.factor(Monthly_Species_Counts$ParkName)
Monthly_Species_Counts$Species <- as.factor(Monthly_Species_Counts$Species)

# Tukey's HSD test for the interaction term (ParkName * Species)
library(agricolae)
monthly_tukey_result <- TukeyHSD(monthly_anova_result, "ParkName:Species")

# Extract significant results
monthly_tukey_df <- as.data.frame(monthly_tukey_result$`ParkName:Species`)
monthly_tukey_df$Comparison <- rownames(monthly_tukey_result$`ParkName:Species`)
rownames(monthly_tukey_df) <- NULL

# Filter significant comparisons
monthly_significant_tukey <- subset(monthly_tukey_df, `p adj` < 0.05)

# Print the top 10 significant results
head(monthly_significant_tukey, 10)
```

>Post-Hoc Result Interpretation:

> In total, the post-hoc test for monthly activity levels revealed 70 significant comparisons, a notably smaller number than the 214 significant comparisons for daily activity levels. These results suggest there are differences in the resolution and variability of the data at these two time scales, with daily data being more granular and sesitive to short-term differences while monthly data reflect broader patterns. 

> The top 10 most significant results are between 1) Mill Creek Park eastern gray squirrel and Bells Bend Park American beaver, 2) Bells Bend Park white-tailed deer and Bells Bend Park American beaver, 3) Mill Creek Park white-tailed deer and Mill Creek Park American beaver, 4) Bells Bend Park white-tailed deer and Mill Creek Park American beaver, 5) Mill Creek Park eastern gray squirrel and Beaman Park bobcat, 6) Bells Bend Park white-tailed deer and Beaman Park bobcat, 7) Mill Creek Park eastern gray squirrel and Bells Bend Park bobcat, 8) Bells Bend Park white-tailed deer and Bells Bend Park bobcat, 9) Mill Creek Park eastern gray squirrel and Beaman Park coyote, and 10) Bells Bend Park white-tailed deer and Beaman Park coyote.

>The Tukey test results on the monthly species activity dataset reveal significant differences in activity levels between species and parks. For instance, the activity of "Eastern Gray Squirrel" in Mill Creek Park was significantly higher compared to "American Beaver" in Bells Bend Park (difference: 50.58, 95% confidence interval: [36.02, 65.14], adjusted p-value: < 0.001). Similarly, "White-tailed Deer" in Bells Bend Park exhibited significantly higher activity compared to "American Beaver" in the same park (difference: 36.93, 95% confidence interval: [22.37, 51.49], adjusted p-value: < 0.001).

>Within Bells Bend Park, "White-tailed Deer" consistently showed higher activity levels compared to other species. For example, their activity was significantly greater than "Bobcat" (difference: 34.83, 95% confidence interval: [20.27, 49.39], adjusted p-value: < 0.001) and "Virginia Opossum" (difference: 36.43, 95% confidence interval: [22.17, 50.99], adjusted p-value: < 0.001). These results indicate that "White-tailed Deer" dominate activity levels within this park.

>Interspecies comparisons across parks also highlighted notable patterns. The activity of "Virginia Opossum" in Bells Bend Park was significantly lower than "Eastern Gray Squirrel" in Mill Creek Park (difference: -47.67, 95% confidence interval: [-62.22, -33.11], adjusted p-value: < 0.001). Similarly, within Mill Creek Park, "White-tailed Deer" showed higher activity compared to "Muskrat" (difference: 36.68, 95% confidence interval: [22.12, 51.24], adjusted p-value: < 0.001).

>Overall, species such as "Eastern Gray Squirrel" in Mill Creek Park and "White-tailed Deer" in Bells Bend Park exhibited the highest activity levels. This may be attributed to favorable environmental conditions or species-specific behaviors in these parks. Conversely, species like "Virginia Opossum" and "American Beaver" consistently exhibited lower activity levels. These findings suggest that ecological or habitat-specific factors significantly influence species activity patterns across parks.

# Two sample T-test

## Daily Species Activity 

```{r}
# Independently filter for three potentially important species: Eastern Gray Squirrel, White-tailed Deer, and Coyote
dmSquirrel <- Daily_Merge %>% 
  filter(Species == "Eastern Gray Squirrel") 
dmDeer <- Daily_Merge %>% 
  filter(Species == "White-tailed Deer") 
dmCoyote <- Daily_Merge %>% 
  filter(Species == "Coyote")
```

### Park Classification (Regional vs Community)
```{r}
# Run a t-test for each species
SQ_t_test_class <- t.test(CountAdjusted ~ Classification, data = dmSquirrel)
SQ_t_test_class

D_t_test_class <- t.test(CountAdjusted ~ Classification, data = dmDeer)
D_t_test_class

C_t_test_class <- t.test(CountAdjusted ~ Classification, data = dmCoyote)
C_t_test_class

#Squirrel and Deer both significant, coyote not. 
```

>T-test Result Interpretation:

>For the daily activity of eastern gray squirrel, community and regional parks have statistically different means (p-value < 2.2e-16). The sample estimate for community parks is about 1.82 while the sample estimate for regional parks is 0.201.

>For the daily activity of white-tailed deer, community and regional parks have statistically different means (p-value < 2.2e-16). The sample estimate for community parks is about 0.176 while the sample estimate for regional parks is 0.853.

>For the daily activity of coyote, community and regional parks have statistically the same means (p-value = 0.1471). The sample estimate for community parks is about 0.0363 while the sample estimate for regional parks is about 0.0575.

### Park Size (>1000ac vs <1000ac)
```{r}
# Run a t-test for each species
SQ_t_test_size <- t.test(CountAdjusted ~ Size, data = dmSquirrel)
SQ_t_test_size

D_t_test_size <- t.test(CountAdjusted ~ Size, data = dmDeer)
D_t_test_size

C_t_test_size <- t.test(CountAdjusted ~ Size, data = dmCoyote)
C_t_test_size
```

>T-test Result Interpretation:

>For the daily activity of eastern gray squirrel, parks less than 1000 acres and parks more than 1000 acres have statistically different means (p-value < 2.2e-16). The sample estimate for parks less than 1000 acres is about 1.00 while the sample estimate for parks more than 1000 acres is 0.215.

>For the daily activity of white-tailed deer, parks less than 1000 acres and parks more than 1000 acres have statistically different means (p-value = 6.312e-07). The sample estimate for parks less than 1000 acres is about 0.754 while the sample estimate for parks more than 1000 acres is 0.374.

>For the daily activity of coyote, parks less than 1000 acres and parks more than 1000 acres have statistically the same means (p-value = 0.1039). The sample estimate for parks less than 1000 acres is about 0.0578 while the sample estimate for parks more than 1000 acres is about 0.0357.

### Disruptive Activity, i.e., mountain biking, camping and fishing (Yes vs No)
```{r}
# Run a t-test for each species
SQ_t_test_disrupt <- t.test(CountAdjusted ~ DisruptiveActivities, data = dmSquirrel)
SQ_t_test_disrupt

D_t_test_disrupt <- t.test(CountAdjusted ~ DisruptiveActivities, data = dmDeer)
D_t_test_disrupt

C_t_test_disrupt <- t.test(CountAdjusted ~ DisruptiveActivities, data = dmCoyote)
C_t_test_disrupt
```

>T-test Result Interpretation:

>For the daily activity of eastern gray squirrel, parks without disruptive activities and parks with disruptive activities have statistically different means (p-value < 2.2e-16). The sample estimate for parks without disruptive activities is about 1.02 while the sample estimate for parks without disruptive activities is 0.187.

>For the daily activity of white-tailed deer, parks without disruptive activities and parks with disruptive activities have statistically different means (p-value < 2.2e-16). The sample estimate for parks without disruptive activities is about 0.275 while the sample estimate for parks with disruptive activities is 1.33.

>For the daily activity of coyote, parks without disruptive activities and parks with disruptive activities have statistically different means (p-value = 0.04813). The sample estimate for parks without disruptive activities is about 0.0360 while the sample estimate for parks with disruptive activities is about 0.0793.

### Monthly species activity
```{r}

mmSquirrel <- Monthly_Merge %>%
  filter(Species == "Eastern Gray Squirrel")
  
mmDeer <- Monthly_Merge %>%
  filter(Species == "White-tailed Deer")
  
mmCoyote <- Monthly_Merge %>%
filter(Species == "Coyote")
```

### Park Classification (Regional vs. Community)
```{r}
SQ_t_test_class2 <- t.test(CountAdjusted ~ Classification, data = mmSquirrel)
SQ_t_test_class2

D_t_test_class2 <- t.test(CountAdjusted ~ Classification, data = mmDeer)
D_t_test_class2

C_t_test_class2 <- t.test(CountAdjusted ~ Classification, data = mmCoyote)
C_t_test_class2
```

>T-test Result Interpretation:

>For the monthly activity of eastern gray squirrel, community and regional parks have statistically different means (p-value = 0.004491). The sample estimate for community parks is about 50.7 while the sample estimate for regional parks is 5.58. 

>For the monthly activity of white-tailed deer, community and regional parks have statistically different means (p-value = 0.001046). The sample estimate for community parks is about 4.89 while the sample estimate for regional parks is 23.7.

>For the monthly activity of coyote, community and regional parks have statistically the same means (p-value = 0.3486). The sample estimate for community parks is about 1.01 while the sample estimate for regional parks is about 1.60.

### Park Size (>1000ac vs <1000ac)
```{r}
SQ_t_test_size2 <- t.test(CountAdjusted ~ Size, data = mmSquirrel)
SQ_t_test_size2

D_t_test_size2 <- t.test(CountAdjusted ~ Size, data = mmDeer)
D_t_test_size2

C_t_test_size2 <- t.test(CountAdjusted ~ Size, data = mmCoyote)
C_t_test_size2
```

>T-test Result Interpretation:

>For the monthly activity of eastern gray squirrel, parks less than 1000 acres and parks more than 1000 acres have statistically different means (p-value = 0.01198). The sample estimate for parks less than 1000 acres is about 27.9 while the sample estimate for parks more than 1000 acres is 5.98.

>For the monthly activity of white-tailed deer, parks less than 1000 acres and parks more than 1000 acres have statistically the same means (p-value = 0.07912). The sample estimate for parks less than 1000 acres is about 21.0 while the sample estimate for parks more than 1000 acres is 10.4.

>For the monthly activity of coyote, parks less than 1000 acres and parks more than 1000 acres have statistically the same means (p-value = 0.252). The sample estimate for parks less than 1000 acres is about 1.61 while the sample estimate for parks more than 1000 acres is about 0.992.

### Disruptive Activity, i.e., mountain biking, camping and fishing (Yes vs No)
```{r}
SQ_t_test_disrupt2 <- t.test(CountAdjusted ~ DisruptiveActivities, data = mmSquirrel)
SQ_t_test_disrupt2

D_t_test_disrupt2 <- t.test(CountAdjusted ~ DisruptiveActivities, data = mmDeer)
D_t_test_disrupt2

C_t_test_disrupt2 <- t.test(CountAdjusted ~ DisruptiveActivities, data = mmCoyote)
C_t_test_disrupt2
```

>T-test Result Interpretation:

>For the monthly activity of eastern gray squirrel, parks without disruptive activities and parks with disruptive activities have statistically different means (p-value = 0.008249). The sample estimate for parks without disruptive activities is about 28.3 while the sample estimate for parks without disruptive activities is 5.19.

>For the monthly activity of white-tailed deer, parks without disruptive activities and parks with disruptive activities have statistically different means (p-value = 0.001368). The sample estimate for parks without disruptive activities is about 7.65 while the sample estimate for parks with disruptive activities is 37.0.

>For the monthly activity of coyote, parks without disruptive activities and parks with disruptive activities have statistically the same means (p-value = 0.1408). The sample estimate for parks without disruptive activities is about 1.00 while the sample estimate for parks with disruptive activities is about 2.20.
